How to improve PCA based methods of genome scan using ecological data: detecting selection using RDA.

### 2017-02-23 article_genome_scan.tex
update

where $\beta_{jk}$ is the regression coefficient corresponding to the j-th SNP regressed by the k-th ordination axis, and $\epsilon_{j}$ is the residuals vector. To summarize the result of the regression analysis for the j-th SNP, we return a vector of z-scores $z_{j} = (z_{j1}, ..., z_{jK})$ where $z_{jk}$ corresponds to the z-score obtained when regressing the j-th SNP by the k-th ordination axis.
The test statistic is a robust Mahalanobis distance D computed using covRob function of the robustR package. D should be Khi2 distributed after a correction with inflation factor (Luu et al., 2016). Pvalues are computed using K degree of freedom. We use the FDR approach to control for false positives. Qvalue are computed with qvalueR package and a loci is considered as an outlier if its qvalue is less than 10\%.
For the analysis of simulated dataset (see below), we retain the first four ordination axis to compute Mahalanobis distances  as they seem to explain most of the variance in the data. To peform the ordination, we use the 10th environmental variables as input in the explanatory matrix. In the following example, we don't use phenotypic informations since these informations are often laking in environmental genomics. Neither we use geographical coordinates $(i,j)$ which is sometimes added to control for the geographical covariation in the differentiation pattern \citep{Frichot2013}.

To emphasize the utility of RDA, we compared to pcadapt from which the idea of using multivariate method for genome scan is based. On the simulated dataset, we retain $K=3$ axis to compute Mahalanobis distances  as it seems to explain the main amount of variance in the data using scatter plots. To control for false positive, we used the same qvalue threshold (i.e. <= 10\%).

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.5 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 10 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.

@@

From the analysis of Chinook Salmon, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, a heat shock protein which are known to be involved in adaptation to temperature or lipolysis-stimulated lipoprotein receptor which are involved in regulation of lipid metabolic process. This latter process can reasonably thought to be involved in adaptation to food abundance and the need for salmon to migrate on a short or long distance.

>>>>

where $\beta_{jk}$ is the regression coefficient corresponding to the j-th SNP regressed by the k-th ordination axis, and $\epsilon_{j}$ is the residuals vector. To summarize the result of the regression analysis for the j-th SNP, we return a vector of z-scores $z_{j} = (z_{j1}, ..., z_{jK})$ where $z_{jk}$ corresponds to the z-score obtained when regressing the j-th SNP by the k-th ordination axis.
The test statistic is a robust Mahalanobis distance D computed using covRob function of the robustR package. We retain $K=5$ axis to compute Mahalanobis distances as it seems to explain most of the variance. D should be Khi2 distributed after a correction with inflation factor (Luu et al., 2016). Pvalues are computed using K degree of freedom. We use the FDR approach to control for false positives. Qvalue are computed with qvalueR package and a loci is considered as an outlier if its qvalue is less than 10\%.
For the analysis of simulated dataset (see below), we retain the first four ordination axis to compute Mahalanobis distances  as they seem to explain most of the variance in the data. To peform the ordination, we use the 10th environmental variables as input in the explanatory matrix. In the following example, we don't use phenotypic informations since these informations are often laking in environmental genomics. Neither we use geographical coordinates $(i,j)$ which is sometimes added to control for the geographical covariation in the differentiation pattern \citep{Frichot2013}.

To emphasize the utility of RDA, we compared to pcadapt from which the idea of using multivariate method for genome scan is based. On the simulated dataset, we retain $K=5$ axis to compute Mahalanobis distances  as it seems to explain the main amount of variance in the data using scatter plots. To control for false positive, we used the same qvalue threshold (i.e. <= 10\%).

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.1 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 10 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.

@@

From the analysis of Chinook Salmon, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, a heat shock protein which are known to be involved in adaptation to temperature or lipolysis-stimulated lipoprotein receptor which are involved in regulation of lipid metabolic process. This latter process can reasonably thought to be involved in adaptation to food abundance and the need for salmon to migrate on a short or long distance.
Outliers picked up by RDA based method are more easily interpretable as this is illustrated by Fig SXXX. This was of course expected since the ordination is constrained by environmental axis. One can identify an outlier to a selective gradient which was used in the inference process whereas outliers picked up by PCA based method have to be interpreted a posteriori and this can be more difficult to justify. It is however noticeable that a substantial proportion of outliers are shared between PCA and RDA based methods which highlight the fact that both approach are very similar. The lack of interpretability of PCA method is compensated by the fact that it is a blind method in regard to environmental data so it can pick up genes that we will miss using RDA because we miss some crucial environmental data. 
In conclusion, this paper shows that both methods are complementary and should be used simultaneously on the same dataset. RDA will be more easily interpretable in terms of mechanism and should pick up more genes when some important selective gradient are identified but PCA might be able to pick up outliers where no environmental data are available. However this latter method will miss outliers when selective gradient are poorly correlated to the population structure (i.e. geographical distance between populations and individuals). It is therefore crucial when performing landscape genomics to clearly characterize the environmental condition by measuring important variable that are suspected to put selective pressures on the species of interest. This can be done through RDA as it has been validated in our simulations and in the real dataset on Chinook Salmon. 
A myriad of constrained ordination method exist that could be used and allows to control for confounding variable (i.e. partial RDA) for instance altitude, latitude and longitude or allow to perform none linear regression between genetic and environmental data (i.e. LVM - Latent Variable Model). This is still to be tested and explored using both simulation and real data set validation.
